Haploid Selection, Sex Ratio Bias, and Transitions Between Sex-Determination Systems

### 2018-06-02 LaTeX/main_final.tex
revision 2 complete by matt

% Extra symbols added
\usepackage{stix}
%Three part table - I think this is allowed
\usepackage[flushleft]{threeparttable}

@@

%reference equations in appendices
\usepackage{xr}
\externaldocument{S1Appendix_revised}
\externaldocument{S2Appendix_revised}
\externaldocument{S3Appendix_revised}

@@

\end{flushleft}
% Please keep the abstract below 300 words
%currently 308 words
\section*{Abstract}
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. %, which is \textcolor{red}{not expected from previous theory}. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
%Furthermore, we find that the selective strengths exerted by the sex-ratio and associations with selected alleles are often equivalent. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
%Furthermore, we find that the selective forces to equalize the sex ratio are equally important to selective forces that generate skewed sex ratios when accounting for the spread of new sex chromosomes.
%This allows sex-ratio biases to increase or decrease with the spread of new sex chromosomes and implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio. 
Overall, our models predict that sex determination systems should be highly dynamic, particularly when haploid selection is present, consistent with the evolutionary lability of this trait in many taxa.
%Overall, our models reveal that transitions between sex-determining systems, particularly transitions where the heterogametic sex changes, can be driven by loci in previously unpredicted genomic locations that experience selection during diploid and/or haploid phases.
%Our results predict conditions under which sex-determining systems are likely to be labile and draw connections with sex-ratio evolution.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%currently 193 words
\section*{Author summary}
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which biases the sex-ratio and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking alongside driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
We also find that new sex-determining alleles can spread despite being less closely linked to selected loci as long as initial linkage is tight or haploid selection is present.
Our models therefore predict that loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

% Use "Eq" instead of "Equation" for equation citations.
\section*{Introduction}
%TRANSITIONS ARE COMMON
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx, Pennell2018}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including %non-monotreme? I think monotremes have XY but tonnes of translocations, giving: X1,X2...X6 Y1
mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes, due either to translocation of the master sex-determining locus or to the evolution of a new master locus.  
During these transitions, the heterogametic sex can remain the same (hereafter `cis-GSD transitions') as in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
%Presentation at evolution found neo sex chromosome in birds, doesn't seem to be published or bioRxiv yet (consider pers com.), title: A previously unknown neo-sex chromosome mediates plumage divergence and speciation in hybridizing birds Jason Sardell; Elizabeth Cooper; J. Albert C. Uy. I wonder if this is actually a fusion event though?
Alternatively, species can switch between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), as in snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc, Pennell2018}.
In sum, accumulating evidence indicates that transitions between sex-determining systems are common~\cite{Bachtrog:2014bx}.

%MAJOR MECHANISMS
%(1) sex-ratio selection
%Predominant theories accounting for transitions between GSD systems by selection involve sex-ratio selection or selective differences between the sexes (e.g., sexually antagonistic selection)~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}. 
%\citet{Bull:1977wt} shows that direct selection on new sex determination loci can allow them to spread and switch between male and female heterogamety. 
It has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
Classic `Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}. 
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex, including new sex determination systems.
%Because the sex ratio is directly determined by the sex-determining system, it has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}).

%MAJOR MECHANISMS
%(2) sex-antagonism
The evolution of sex determination is also thought to be strongly influenced by differences in selection between the sexes ~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}.
For example, loci experiencing sexual antagonism have been shown to favour the spread of new genetic sex-determining alleles that are closely linked~\cite{vanDoorn:2007eu,vanDoorn:2010hu,Muralidhar2018}. 
%van Doorn and Kirkpatrick~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and Muralidhar and Veller~\cite{Muralidhar2018} have shown that new sex-determining alleles can be favoured if they arise at loci in close linkage with a locus that experiences sexual antagonism. 
Linkage allows a stronger favourable association to build up between a male-beneficial allele and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles in a population with ESD~\cite{Muralidhar2018}.
%However, any sexually-antagonistic loci that are more closely linked to the ancestral sex-determining locus should develop similar, favourable associations and are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}. 
By similar logic, however, existing sexually-antagonistic alleles associated with the current sex-determining locus are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.

%Tight linkage Model Feature and Result
One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome). %during trans-GSD transitions
%, when considering loci in very tight linkage with the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome), 
%\textcolor{red}{[vDK2007 does "explicitly" consider this as well (although on pg 15 of their SOM they say they will ignore equilibrium A, as it is stable under a "limited range of biologically relevant parameter combinations"; perhaps haploid selection changes this impression), but only for cis-GSD transitions. They find that if equil B is stable the neo-Y never spreads when r=0. vDK2010 does not "explicitly" calculate pX and pY, which are required to see if their equation 7 is positive or negative, but nowhere do they seem to say that weaker linkage \textit{can't} evolve given r initially small.]}
%we explicitly calculate ancestral polymorphisms. 
%Our analysis reveals that, under various forms of selection, the ancestral-X can carry suboptimal alleles for females and/or the ancestral-Y can carry female-beneficial alleles. 
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
In addition, selection on ancestral-X chromosomes in males can prevent the X from becoming optimally specialised for female-beneficial alleles. 
These factors, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which was not revealed by previous theory \cite{vanDoorn:2010hu}. A similar argument applies to ZW$\rightarrow$XY transitions. 
Thus, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

%Haploid selection model feature and discussion of why needed
Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for loci that experience diploid-sex-differences in selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

%Previous models with haploid selection
Two previous studies have considered the spread of GSD with sex-limited meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles can overwhelm selection to balance the sex ratio. %, at least when new sex-determining alleles are rare. 
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver and therefore exhibiting an ancestral sex ratio bias. 
They found that a new, completely unlinked, GSD system can spread if it generates the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection can overwhelm the benefits of being associated with driving alleles. 
It is thus currently unclear when haploid selection favors increased versus decreased linkage between haploid selected loci and a new sex-determining locus. 
In addition, because the sex ratio is determined by linkage between haploid selected loci and the sex-determining locus, it is also unclear when Fisherian sex-ratio selection is the most important driver of transitions between sex-determining systems.

%main results
Here, we analytically find the conditions under which new GSD or ESD systems spread in ancestral GSD systems with any degree of linkage between the loci involved and arbitrary forms of haploid and diploid selection. 
Doing so, we reconcile and generalize the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by deriving conditions for the spread of new GSD systems that alter linkage with haploid selected loci.
Our result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
We also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that transitions in sex-determining system are favoured \textit{equally strongly} in situations where sex-ratio biases increase or decrease (and in situations where sex-ratio biases are ancestrally present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which is not predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
%Together, our results suggest that neither selection to equalize the sex ratio nor the benefits of associating with haploid selected alleles is an overwhelming force driving transitions between sex-determining systems. 
Together, our results suggest that both selection to equalize the sex ratio and the benefits of associating with haploid selected alleles can drive transitions between sex-determining systems, leading to stronger or weaker sex-linkage and increased or decreased sex ratio bias.

%Here we use mathematical models to find the conditions under which new sex-determining systems spread when individuals experience selection at both diploid and haploid stages, which allows fitness differences between the sexes and sex ratio biases to occur simultaneously

%Haploid selection results

%The sex ratio is directly determined by the sex-determining system, and it has therefore been suggested that sex-ratio selection is a dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
%`Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}.
%This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
%Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
%Selection therefore favours mutants that increase investment in the rarer sex.

%The default mode of sex-ratio selection is `Fisherian' sex-ratio selection, which favours equal investment in male and female offspring \citep[i.e., a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce,][]{Fisher:1930wy,Charnov:1982wg}. 
%Given that the sex-determining system can directly affect the sex ratio, we might expect Fisherian sex-ratio selection to influence the spread of new sex-determining systems. 
%In the case of sex-chromosome evolution, Kozielska et al. (2010)~\cite{Kozielska:2010vm} consider systems in which the ancestral sex chromosomes experience meiotic drive (e.g., where driving X or Y chromosomes are inherited disproportionately often), which causes sex ratios to become biased~\cite{Hamilton:1967ts}. 
%They find that masculinizing or feminizing mutations that arise at a new, unlinked locus (i.e., neo-Y or neo-W chromosomes) can then spread, which restore an even sex ratio. 
%However, starting from ESD, Ubeda et al. (2015)~\cite{Ubeda:2015fx} have numerically shown that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread (i.e., ESD $\rightarrow$ GSD) despite causing a sex-ratio bias.

%HOLES IN CURRENT THEORY
%(1) haploid selection; it exists in many spp
%While both opposing selection in the diploid sexes and sex-ratio selection surely influence the evolution of new sex-determining factors, so too might selection in the haploid phase.

%HOLES IN CURRENT THEORY
%(2) haploid selection; it can complicate outcomes for SD transitions
%There are various ways by which haploid selection could influence transitions between sex-determining systems. 
%If we assume that haploid selection at any particular locus predominantly occurs in one sex (e.g., meiotic drive during spermatogenesis), then such loci experience a form of sex-specific selection. 
%In this respect, we might expect that haploid selection would affect transitions between sex-determining systems in a similar manner to sex-specific diploid selection (as explored in~\cite{vanDoorn:2007eu,vanDoorn:2010hu}). 
%That is, new masculinizing mutations (neo-Y alleles) could be favoured via associations with alleles that are beneficial in the male haploid stage (e.g., \cite{Ubeda:2015fx}). 
%Kozielska et al. (2010)~\cite{Kozielska:2010vm} have shown this to be theoretically possible for the special case of perfect linkage between the new sex-determining locus and a locus that experiences meiotic drive in one sex, under a limited number of forms of diploid selection.
%Similarly, Ubeda et al. (2015)~\cite{Ubeda:2015fx} numerically show that sex-specific meiotic drive can allow masculinizing and feminizing mutations to spread in an ancestrally environmental-sex-determination system, again under a limited number of forms of diploid selection. 
%On the other hand, sex ratios can also become biased by linkage between the sex-determining locus and a locus that harbours genetic variation in haploid fitness, as indeed occurs in the two examples above.
%There are also several known empirical cases of sex-ratio bias caused by sex-linked meiotic drive alleles (Burt and Trivers, 2006, Chapter 6~\cite{Burt:2006}) or selection among X- and Y-bearing pollen~\cite{Lloyd:1974tz,Conn:1981uw,Stehlik:2005ul,Stehlik:2006to,Field:2012fd,Field:2013cc}. 
%Sex-ratio selection can thus select against the spread of linked sex-determining alleles and favour the invasion of alternative, unlinked sex-determination systems \cite{Kozielska:2010vm,Ubeda:2015fx}.
%It is not yet clear when new sex-determining alleles are favoured by associations with haploid selected alleles vs. sex-ratio bias, especially under a wide range of possible diploid selection regimes (e.g., sex-antagonism, overdominance).

>>>>

% Extra symbols added (male and female)
\usepackage{stix}
%Three part table 
\usepackage[flushleft]{threeparttable}

@@

%reference equations in appendices
%\usepackage{xr}
%\externaldocument{S1Appendix_revised}
%\externaldocument{S2Appendix_revised}
%\externaldocument{S3Appendix_revised}

@@

\end{flushleft}
% Please keep the abstract below 300 words
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Sex determination is remarkably dynamic; many taxa display shifts in the location of sex-determining loci or the evolution of entirely new sex-determining systems. 
Predominant theories for why we observe such transitions generally conclude that novel sex-determining systems are favoured by selection if they equalise the sex ratio or increase linkage with a locus that experiences different selection in males vs.\ females. 
We use population genetic models to extend these theories in two ways: 
(1) We consider the dynamics of loci very tightly linked to the ancestral sex-determining loci, e.g., within the non-recombining region of the ancestral sex chromosomes. 
Variation at such loci can favour the spread of new sex-determining systems in which the heterogametic sex changes (XY to ZW or ZW to XY) and the new sex-determining region is less closely linked (or even unlinked) to the locus under selection. 
(2) We consider selection upon haploid genotypes either during gametic competition (e.g., pollen competition) or meiosis (i.e., non-Mendelian segregation), which can cause the zygotic sex ratio to become biased. 
Haploid selection can drive transitions between sex-determining systems without requiring selection to act differently in diploid males vs.\ females. 
With haploid selection, we find that transitions between male and female heterogamety can evolve where linkage with the sex-determining locus is either strengthened or weakened. 
Furthermore, we find that sex-ratio biases may increase or decrease with the spread of new sex chromosomes, which implies that transitions between sex-determining systems cannot be simply predicted by selection to equalise the sex ratio.
In fact, under many conditions, we find that transitions in sex determination are favoured equally strongly in cases where the sex ratio bias increases or decreases.
Overall, our models predict that sex determination systems should be highly dynamic, particularly when haploid selection is present, consistent with the evolutionary lability of this trait in many taxa.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Author summary}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Systems of sex determination are strikingly diverse and labile in many clades. 
This poses the question: what drives transitions between sex-determining systems? 
Here, we use models to derive conditions under which new sex-determining systems spread. 
Prevailing views suggest that new sex-determining systems are favoured when they equalize the sex ratio and/or when they are more closely linked to genes that experience differential selection in males and females. 
Our models include selection upon haploid genotypes (meiotic drive or gametic competition), which biases the sex-ratio and occurs differently in male and female gametes. 
Surprisingly, we find the two forces (selection to equalize the sex ratio and the benefits of hitchhiking alongside driven alleles that distort the sex ratio) will often be equally strong, and thus neither is sufficient to explain the spread of new sex-determining systems in every case. 
We also find that new sex-determining alleles can spread despite being less closely linked to selected loci as long as initial linkage is tight or haploid selection is present.
Our models therefore predict that loci in previously unexpected genomic locations and/or experiencing various types of selection (including haploid selection) can now be implicated as drivers of transitions between sex-determining systems.

@@

% Use "Eq" instead of "Equation" for equation citations.
%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%
Animals and angiosperms exhibit extremely diverse sex-determining systems (reviewed in~\cite{Bull:1983vi,Charlesworth:2010it,Beukeboom:2014vb,Bachtrog:2014bx, Pennell2018}). 
Among species with genetic sex determination (GSD), some taxa have heterogametic males (XY) and homogametic females (XX), including mammals and most dioecious plants~\cite{Ming:2011iy}; whereas other taxa have homogametic males (ZZ) and heterogametic females (ZW), including Lepidoptera and birds. 
Within several taxa, the chromosome that harbours the master sex-determining locus changes, due either to translocation of the master sex-determining locus or to the evolution of a new master locus.  
During these transitions, the heterogametic sex can remain the same (hereafter `cis-GSD transitions') as in Salmonids~\cite{Li:2011fm,Yano:2012di}, Diptera~\cite{Vicoso:2015hf}, and \textit{Oryzias}~\cite{Myosho:2012fv}. 
Alternatively, species can switch between male and female heterogamety (XY$\leftrightarrow$ZW, hereafter `trans-GSD transitions'), as in snakes~\cite{Gamble2017}, lizards~\cite{Ezaz:2009tk}, eight of 26 teleost fish families~\cite{Mank:2006bt}, true fruit flies (Tephritids,~\cite{Vicoso:2015hf}), amphibians~\cite{Hillis:1990gu}, the angiosperm genus \textit{Silene}~\cite{Slancarova:2013dq}, the angiosperm family \textit{Salicaceae}~\cite{Pucholt2015,Pucholt2017} and Coleoptera and Hemiptera (plate 2~\cite{Beukeboom:2014vb}).
Indeed, in some cases, both male and female heterogametic sex-determining systems can be found in the same species, as reported in houseflies~\cite{Macdonald1978}, midges~\cite{Thompson1971}, frogs~\cite{Ogata:2007jm}, cichlid fish~\cite{Ser:2010iq}, tilapia~\cite{Lee2004}, sea bass~\cite{Vandeputte2007}, and lab-strains of Zebrafish~\cite{Liew2012,Wilson2014}.
In addition, multiple transitions have occurred between genetic and environmental sex-determining systems (GSD$\leftrightarrow$ESD), e.g., in reptiles and fishes~\cite{Conover:1987in,Mank:2006bt,Pokorna:2009ui,Ezaz:2009tk,Pen:2010kk,Holleley:2015hc, Pennell2018}.
In sum, accumulating evidence indicates that transitions between sex-determining systems are common~\cite{Bachtrog:2014bx}.

It has been suggested that sex-ratio selection is a particularly dominant force in the evolution of sex determination (e.g., Bull, 1983, p 66-67 ~\cite{Bull:1983vi}; Buekeboom and Perrin, 2014, Chapter 7~\cite{Beukeboom:2014vb}). 
Classic `Fisherian' sex-ratio selection favours a 1:1 zygotic sex ratio when assuming that males and females are equally costly to produce~\cite{Fisher:1930wy,Charnov:1982wg}. 
This follows from the fact that, for an autosomal locus, half of the genetic material is inherited from a male and half from a female~\cite{West:2009we}. 
Thus, if the sex ratio is biased, an individual of the rarer sex will, on average, contribute more genetic material to the next generation. 
Selection therefore typically favours mutants that increase investment in the rarer sex, including new sex determination systems.

The evolution of sex determination is also thought to be strongly influenced by differences in selection between the sexes ~\cite{Blaser2012, Beukeboom:2014vb,vanDoorn2014re}.
For example, loci experiencing sexual antagonism have been shown to favour the spread of new genetic sex-determining alleles that are closely linked~\cite{vanDoorn:2007eu,vanDoorn:2010hu,Muralidhar2018}. 
Linkage allows a stronger favourable association to build up between a male-beneficial allele and a neo-Y allele, for example. 
Such associations can favour cis-GSD transitions~\cite{vanDoorn:2007eu}, trans-GSD transitions~\cite{vanDoorn:2010hu}, and new partially-masculinizing or partially-feminizing alleles in a population with ESD~\cite{Muralidhar2018}.
By similar logic, however, existing sexually-antagonistic alleles associated with the current sex-determining locus are expected to hinder the spread of a new sex-determining system~\cite{vanDoorn:2007eu,vanDoorn:2010hu}.

One novel feature of the models developed here is that we explicitly consider the maintenance of genetic variation around the ancestral sex-determining locus (e.g., within the non-recombining region of a sex chromosome).  
Counterintuitively, when linkage is tight between the sex-determining locus and a selected locus, an allele good for females can be at higher frequency on the ancestral-Y than on the ancestral-X under a variety of forms of selection.
In addition, selection on ancestral-X chromosomes in males can prevent the X from becoming optimally specialised for female-beneficial alleles. 
These factors, in turn, can favour a new ZW sex-determining locus that has weaker linkage with loci under selection, which was not revealed by previous theory \cite{vanDoorn:2010hu}. A similar argument applies to ZW$\rightarrow$XY transitions. 
Thus, we show that selected loci in very tight linkage with the ancestral GSD locus can favour trans-GSD transitions during which linkage associations are actually weakened.

Most significantly, we include haploid selection (gametic competition or meiotic drive) in models describing cis-GSD, trans-GSD, and GSD to ESD transitions.
This poses an apparent evolutionary problem. 
On one hand, haploid selection is typically sex-limited in that it usually occurs among gametes produced by one sex only~\cite{Mulcahy:1996ha,JOSEPH:2004haa,Ubeda:2005gw,Lindholm:2016cw}. 
Therefore, one might expect new sex-determining systems to benefit from close linkage with haploid selected loci, as found for loci that experience diploid-sex-differences in selection~\cite{vanDoorn:2007eu,vanDoorn:2010hu, Muralidhar2018}. 
On the other hand, associations between sex-determining loci and haploid selected loci generate biased zygotic sex ratios, which should generally hinder the spread of new sex-determining systems.

Two previous studies have considered the spread of GSD with sex-limited meiotic drive~\cite{Kozielska:2010vm,Ubeda:2015fx} under a limited number of possible genetic architectures and diploid selective regimes. 
Ubeda et al. (2015)~\cite{Ubeda:2015fx} considered ancestral-ESD (with no sex-ratio bias) and numerically showed that new GSD alleles can spread if they arise in linkage with meiotic drive loci. 
For example, a masculinizing allele spreads in association with an allele that is favoured during male meiosis, causing sex ratios to become male-biased.
This suggests that the benefits of associating with driving alleles can overwhelm selection to balance the sex ratio.
However, Kozielska et al. (2010)~\cite{Kozielska:2010vm} considered an ancestral GSD system that is perfectly linked to a meiotic driver and therefore exhibiting an ancestral sex ratio bias. 
They found that a new, completely unlinked, GSD system can spread if it generates the rarer sex, creating a balanced sex ratio. 
This suggests that Fisherian sex-ratio selection can overwhelm the benefits of being associated with driving alleles. 
It is thus currently unclear when haploid selection favors increased versus decreased linkage between haploid selected loci and a new sex-determining locus. 
In addition, because the sex ratio is determined by linkage between haploid selected loci and the sex-determining locus, it is also unclear when Fisherian sex-ratio selection is the most important driver of transitions between sex-determining systems.

Here, we analytically find the conditions under which new GSD or ESD systems spread in ancestral GSD systems with any degree of linkage between the loci involved and arbitrary forms of haploid and diploid selection. 
Doing so, we reconcile and generalize the results of Kozielska et al. (2010)~\cite{Kozielska:2010vm} and Ubeda et al. (2015)~\cite{Ubeda:2015fx} by deriving conditions for the spread of new GSD systems that alter linkage with haploid selected loci.
Our result is qualitatively distinct from those for diploid selection alone~\cite{vanDoorn:2007eu,vanDoorn:2010hu} and suggests that haploid selection is more likely to promote transitions between sex-determination systems. 
We also show that transitions involving haploid selection cannot be simply explained by invoking sex-ratio selection. 
In particular, under a wide range of conditions, we show that transitions in sex-determining system are favoured \textit{equally strongly} in situations where sex-ratio biases increase or decrease (and in situations where sex-ratio biases are ancestrally present or absent). 
Finally, we show that ESD may not evolve, even if the sex ratio is initially biased by haploid selection, which is not predicted by previous theories for transitions to ESD~\cite{Charnov:1982wg,Bull:1983vi,West:2009we}. 
Together, our results suggest that both selection to equalize the sex ratio and the benefits of associating with haploid selected alleles can drive transitions between sex-determining systems, leading to stronger or weaker sex-linkage and increased or decreased sex ratio bias.

%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Model}
%%%%%%%%%%%%%%%%%%%%%%%%%%

We consider transitions between ancestral and novel sex-determining systems using a three-locus model, each locus having two alleles (Fig \ref{fig:model_outline}). A full description of our model, including recursion equations, is given in \nameref{app:recurs}. 
Locus $\mathbf{X}$ is the ancestral sex-determining region, with alleles X and Y (or Z and W).
Locus $\mathbf{A}$ is a locus under selection, with alleles $A$ and $a$.
Locus $\mathbf{M}$ is a novel sex-determining region, at which the null allele ($M$) is initially fixed in the population such that sex of zygotes is determined by the genotype at the ancestral sex-determining region, $\mathbf{X}$; XX genotypes become females and XY become males (or ZW become females and ZZ become males). 
To evaluate the evolution of new sex-determining systems, we consider the spread of a novel sex-determining allele ($m$) at the $\mathbf{M}$ locus.

Here, we assume that the $\mathbf{M}$ locus is `epistatically dominant' over the $\mathbf{X}$ locus such that zygotes with at least one $m$ allele develop as females with probability $k$ and as males with probability $1-k$, regardless of the $\mathbf{X}$ locus genotype. 
With $k=0$, the $m$ allele is a masculinizer (a neo-Y allele) and with $k=1$ the $m$ allele is a feminizer (a neo-W allele). 
With intermediate $k$, we can interpret $m$ as an environmental sex-determination (ESD) allele, such that zygotes develop as females in a proportion ($k$) of the environments they experience. 
The assumption that derived sex-determining loci are epistatically dominant is motivated by empirical systems in which multiple sex determining alleles segregate (i.e., X, Y, Z, and W alleles present), such as, cichlid fish~\cite{Ser:2010iq}, platyfish (\textit{Xiphophorus maculatus}~\cite{Kallman1968}), houseflies (\textit{Musca domestica}~\cite{Dubendorfer2002}), western clawed frogs (\textit{Xenopus tropicalis}~\cite{Roco2015}) and \textit{Rana rugosa}~\cite{Ogata:2007jm}.  
Nevertheless, our supplementary analysis file (\nameref{file:Mathematica}) allows other dominance relationships between loci to be specified (see also \cite{vanDoorn:2007eu} supplementary material for a numerical analysis).

\begin{figure}[!h]
%\centering
%\centerline{\includegraphics[width=0.9\linewidth]{../Plots/Sex_determination_outline.pdf}}
\caption{
{\bf Outline of model features.}
Panel A: Recombination rate parameters between the ancestral-sex-determining locus ($\mathbf{X}$, here assumed to have alleles X and Y), a locus under selection ($\mathbf{A}$, with alleles $A$ and $a$), and a new sex-determining locus ($\mathbf{M}$, with alleles $M$ and $m$). 
Panel B: Haploid selection is often sex-limited, occurring during haploid production or competition in one sex (shown here in males by dashed circles). 
If X or Y alleles are linked with alleles that experience haploid selection in males ($r<1/2$), then zygotic sex ratios can become biased because either X- or Y-bearing male gametes/gametophytes will be more abundant after haploid selection. 
Similarly, zygotic sex-ratio biases can arise if haploid selected alleles are linked with new sex-determining alleles ($R<1/2$). 
However, the zygotic sex ratio is not biased by male haploid selection in ZW sex-determining systems. 
Panel C: During cis-GSD transitions (XY to XY or ZW to ZW), a neo-Y allele ($m$) spreads to pseudo-fixation (i.e, all males bear the neo-Y) and the ancestral Y allele is lost. 
Panel D: During trans-GSD transitions (XY to ZW or ZW to XY), a neo-W allele ($m$) spreads to pseudo-fixation (i.e, all females bear the neo-W) and the ancestral X allele is lost. 
Neo-W alleles allow Y-associated alleles into females, which may impede or aid their spread. 
}
\label{fig:model_outline}
\end{figure}
